Do larger catchments respond different to forest cover change? Re-analysing a global data set.

### 2021-05-30 paper/Forest_and_Water/Forest_and_Water.Rmd
paper intro & methods and new watersheds

The staerting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this dataset

>>>>

## The original data set

The starting point of this paper is the data base of studies which were included in @zhang2017 as supplementary material. The columns in this data set are the watershed number, the	watershed name, the	Area in km^2^, the annual average precipitation	(Pa) in mm, the	forest type, hydrological regime, and climate type, the change in forest cover	in % ($\Delta$F%) and the change in streamflow in % ($\Delta$Qf(%), based on equation 1 in @zhang2017), the precipitation data type, the assessment technique, and the	source of the info, which is a citation. 
Several of these columns contain abbreviations to describe the different variables, which are summarised in Table 1.

Table 1  Summary of abbreviations of factors used in the @zhang2017 data set

|Factor | Abbreviation | Definition |
|:------------|:-------|:----------|
| forest type | CF |coniferous forest | 
|  | BF | broadleaf forest |
|  | MF |mixed forest  |
|hydrological regime|RD|rain dominated|
| |SD|snow dominated|
| climate type| EL| energy limited|
| |WL| water limited|
| |EQ| equitant|
|precipitation data type|OB|observed|
| |SG|spatial gridded|
| |MD|modelled|
|assessment technique|PWE|paired watershed experiment|
| |QPW|quasi-paired watershed experiment|
| |HM|hydrological modelling|
| |EA|elastictity analysis|
| |SH|combined use of statistical methods|
| | |and hydrographs|

While the @zhang2017 use the dryness index in their analysis, potential or reference evapotranspiration is not part of the data set.
We combined the tables for small (< 1000 km^2^) and large (>= 1000 km^2^) watreshed data sets in our analysis. Some small naming errors and citations for some of the data sets for some of the small watersheds were fixed as we were familiar with the studies. But overall the original data set was not changed.

## Additional data collection

To enhance the existing data set, this study added additional variables. The first variables added were the latitude and longitude for the center of the watershed as an approximation of its spatial location. Using this information  annual average reference evapotranspiration (E0) was extracted from **XXXXX** if value of E0 were not available from the original papers. For large water watersheds, this value, similar to annual average rainfall, is an approximation of the climate at the location.

The length of the study can be a variable influencing the change in flow [e.g. @jackson2005] and therefore, the length of the different studies was extracted from the references provided by @zhang2017.

Several additional data points from watershed studies were extracted from @zhang2011, @zhao2010, @borg1988, @thornton2007, @zhou2010, @rodriguez2010, @ruprechtetal1991 and @pena-arancibia2012, and these were checked against the existing studies to prevent overlap. In the citation column in the data set, in general the main reference for the calculated change in streamflow  was used, because sometimes the original study did not provide the quantification of the change in streamflow (i.e. Table 6 in @zhang2011)

The final column in the improved data set is a "notes" column, which is not further used in the analysis, but gives context to some of the data for future research.

## Statistical modelling

To estimate how the change in streamflow is affected by the change in forest cover while considering the effects of the other variables, we applied generalised additive modelling (GAM) [@wood2006]. In particular we fitted the following initial model to replicate the variables in the analysis from @zhang2017:

$$\tag{1}
\begin{aligned}
\Delta \%Q \sim ~&\Delta \%forest + s(Pa) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}$$

However, the overall skewed distribution of the predictant ($\Delta \%Q$) is problematic, and this results in a skewed distribution of the GAM model residuals. As a result we transformed $\Delta \%Q$ and $\Delta \%forest$ back to fractions (0 - 1) and log transformed using $log10(x + 1)$, where $x$ is either $\Delta Q$ or $\Delta forest$. This means that the model residuals are $\sim N(0,\sigma^2)$ results in the following equation:

$$\tag{2}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}$$

In this model, the assumption is that all continuous variables (such as Pa) can have a linear or non-linear relationship with $log10(\Delta Q)$. This means that a smooth function $s()$ is applied to the variable. To restrict the smoothness of the fit, the smoothness factor $k$ is restricted to a value of 3 [@wood2006]. This restriction was applied to smooth variables throughout this paper and we have dropped this from the subsequent equations.

For the model in equation 2, we only used the data from @zhang2017 to make sure that the additional watersheds added to the data set did not influence the analysis. Given that in @zhang2017, dryness ($\frac{E0}{Pa}$) is used to look at variations in the change in flow, we also fitted the following model:

$$\tag{3}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}$$

Subsequently, using the full data set, including the additional watersheds and the additional variables the following two models were fitted:

$$\tag{4}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
\end{aligned}$$

$$\tag{5}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
\end{aligned}$$

The results were analysed to identify:  
1. the significance of the different variables  
2. the direction of the categorical or shape of the smooth variables

### 2021-05-30 paper/Forest_and_Water/Forest_and_Water.tex
paper intro & methods and new watersheds

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Short Paper},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Short Paper}
    \author[The University of Sydney, INIA]{R. Willem Vervoort\corref{1}}
   \ead{willem.vervoort@sydney.edu.au} 
    \author[INIA]{Eliana Nervi}
   \ead{eliananervi@gmail.com} 
    \author[IMFIA]{Jimena Alonso\corref{2}}
   \ead{jalonso@fing.edu.uy} 
      \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney,
Sydney, NSW 2006, Australia}
    \address[INIA]{INIA, Uruguay}
    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of
Engineering, Universidad de la República, 11200 Montevideo, Departamento
de Montevideo, Uruguay}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

@@

\emph{Text based on elsarticle sample manuscript, see
\url{http://www.elsevier.com/author-schemas/latex-instructions\#elsarticle}}

@@

There have been several review papers aiming to summarize different
studies across the globe, in relation to paired watershed studies (Bosch
and Hewlett, 1982; Brown et al., 2005) and more generally (Jackson et
al., 2005; Zhang et al., 2017). These studies are aiming to generalize
the individual findings and to identify if there are global trends or
relationships that can be developed. The most recent review (Zhang et
al., 2017) developed an impressive database of watershed studies in
relation to changes in streamflow due to changes in forest cover based
on a global data set. This dataset, that covers over 250 studies are
described in terms of the change in streamflow as a result of the change
in forest cover, where studies related to both forestation (increase in
forest cover) and deforestation (decrease in forest cover) were
included.

@@

In particular, the objective of this paper is to 1) enhance the data set
from ({\textbf{???}}) with further watersheds and spatial coordinates
and 2) to analyse the possibility of non-linear and partial effects of
the different factors and variables in the data base using generalised
linear (GLM) and generalised additive models (GAM Wood (2006)). Finally
we hope to point to further research that can expand our work and that
outlined Zhang et al. (2017) to better understand the impact of forest
cover change on streamflow.

\hypertarget{front-matter}{%
\section{Front matter}\label{front-matter}}

The author names and affiliations could be formatted in two ways:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\item
  Group the authors per affiliation.
\item
  Use footnotes to indicate the affiliations.
\end{enumerate}

See the front matter of this document for examples. You are recommended
to conform your choice to the journal you are submitting to.

@@

@@

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{longtable}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Do larger catchments respond different to forest cover change? Re-analysing a global data set.},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

  \title{Do larger catchments respond different to forest cover change?
Re-analysing a global data set.}
    \author[The University of Sydney, INIA]{R. Willem Vervoort\corref{1}}
   \ead{willem.vervoort@sydney.edu.au} 
    \author[INIA]{Eliana Nervi}
   \ead{eliananervi@gmail.com} 
    \author[IMFIA]{Jimena Alonso\corref{2}}
   \ead{jalonso@fing.edu.uy} 
      \address[The University of Sydney]{School of Life and Environmental Sciences, The University of Sydney,
Sydney, NSW 2006, Australia}
    \address[INIA]{INIA, Uruguay}
    \address[IMFIA]{Institute of Fluid Mechanics and Environmental Engineering, School of
Engineering, Universidad de la República, 11200 Montevideo, Departamento
de Montevideo, Uruguay}
      \cortext[1]{Corresponding Author}
    \cortext[2]{Equal contribution}

@@

@@

There have been several review papers aiming to summarize different
studies across the globe, in relation to paired watershed studies (Bosch
and Hewlett, 1982; Brown et al., 2005) and more generally (Jackson et
al., 2005; Zhang et al., 2017). These studies are aiming to generalize
the individual findings and to identify if there are global trends or
relationships that can be developed. The most recent review (Zhang et
al., 2017) developed an impressive database of watershed studies in
relation to changes in streamflow due to changes in forest cover based
on a global data set. This dataset, which covers over 250 studies are
described in terms of the change in streamflow as a result of the change
in forest cover, where studies related to both forestation (increase in
forest cover) and deforestation (decrease in forest cover) were
included.

@@

In particular, the objective of this paper is to 1) enhance the data set
from Zhang et al. (2017) with further watersheds and spatial coordinates
and 2) to analyse the possibility of non-linear and partial effects of
the different factors and variables in the data base using generalised
linear (GLM) and generalised additive models (GAM Wood (2006)). Finally
we hope to point to further research that can expand our work and that
outlined Zhang et al. (2017) to better understand the impact of forest
cover change on streamflow.

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\hypertarget{the-original-data-set}{%
\subsection{The original data set}\label{the-original-data-set}}

The starting point of this paper is the data base of studies which were
included in Zhang et al. (2017) as supplementary material. The columns
in this data set are the watershed number, the watershed name, the Area
in km\textsuperscript{2}, the annual average precipitation (Pa) in mm,
the forest type, hydrological regime, and climate type, the change in
forest cover in \% (\(\Delta\)F\%) and the change in streamflow in \%
(\(\Delta\)Qf(\%), based on equation 1 in Zhang et al. (2017)), the
precipitation data type, the assessment technique, and the source of the
info, which is a citation. Several of these columns contain
abbreviations to describe the different variables, which are summarised
in Table 1.

Table 1 Summary of abbreviations of factors used in the Zhang et al.
(2017) data set

\begin{longtable}[]{@{}lll@{}}
\toprule
Factor & Abbreviation & Definition\tabularnewline
\midrule
\endhead
forest type & CF & coniferous forest\tabularnewline
& BF & broadleaf forest\tabularnewline
& MF & mixed forest\tabularnewline
hydrological regime & RD & rain dominated\tabularnewline
& SD & snow dominated\tabularnewline
climate type & EL & energy limited\tabularnewline
& WL & water limited\tabularnewline
& EQ & equitant\tabularnewline
precipitation data type & OB & observed\tabularnewline
& SG & spatial gridded\tabularnewline
& MD & modelled\tabularnewline
assessment technique & PWE & paired watershed experiment\tabularnewline
& QPW & quasi-paired watershed experiment\tabularnewline
& HM & hydrological modelling\tabularnewline
& EA & elastictity analysis\tabularnewline
& SH & combined use of statistical methods\tabularnewline
& & and hydrographs\tabularnewline
\bottomrule
\end{longtable}

While the Zhang et al. (2017) use the dryness index in their analysis,
potential or reference evapotranspiration is not part of the data set.
We combined the tables for small (\textless{} 1000
km\textsuperscript{2}) and large (\textgreater= 1000
km\textsuperscript{2}) watreshed data sets in our analysis. Some small
naming errors and citations for some of the data sets for some of the
small watersheds were fixed as we were familiar with the studies. But
overall the original data set was not changed.

\hypertarget{additional-data-collection}{%
\subsection{Additional data
collection}\label{additional-data-collection}}

To enhance the existing data set, this study added additional variables.
The first variables added were the latitude and longitude for the center
of the watershed as an approximation of its spatial location. Using this
information annual average reference evapotranspiration (E0) was
extracted from \textbf{XXXXX} if value of E0 were not available from the
original papers. For large water watersheds, this value, similar to
annual average rainfall, is an approximation of the climate at the
location.

The length of the study can be a variable influencing the change in flow
(e.g.~Jackson et al., 2005) and therefore, the length of the different
studies was extracted from the references provided by Zhang et al.
(2017).

Several additional data points from watershed studies were extracted
from Zhang et al. (2011), Zhao et al. (2010), Borg et al. (1988),
Thornton et al. (2007), Zhou et al. (2010), Rodriguez et al. (2010),
Ruprecht et al. (1991) and Peña-Arancibia et al. (2012), and these were
checked against the existing studies to prevent overlap. In the citation
column in the data set, in general the main reference for the calculated
change in streamflow was used, because sometimes the original study did
not provide the quantification of the change in streamflow (i.e.~Table 6
in Zhang et al. (2011))

The final column in the improved data set is a ``notes'' column, which
is not further used in the analysis, but gives context to some of the
data for future research.

\hypertarget{statistical-modelling}{%
\subsection{Statistical modelling}\label{statistical-modelling}}

To estimate how the change in streamflow is affected by the change in
forest cover while considering the effects of the other variables, we
applied generalised additive modelling (GAM) (Wood, 2006). In particular
we fitted the following initial model to replicate the variables in the
analysis from Zhang et al. (2017):

\[\tag{1}
\begin{aligned}
\Delta \%Q \sim ~&\Delta \%forest + s(Pa) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

However, the overall skewed distribution of the predictant
(\(\Delta \%Q\)) is problematic, and this results in a skewed
distribution of the GAM model residuals. As a result we transformed
\(\Delta \%Q\) and \(\Delta \%forest\) back to fractions (0 - 1) and log
transformed using \(log10(x + 1)\), where \(x\) is either \(\Delta Q\)
or \(\Delta forest\). This means that the model residuals are
\(\sim N(0,\sigma^2)\) results in the following equation:

\[\tag{2}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa, k = 3) + s(Area, k = 3) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

In this model, the assumption is that all continuous variables (such as
Pa) can have a linear or non-linear relationship with
\(log10(\Delta Q)\). This means that a smooth function \(s()\) is
applied to the variable. To restrict the smoothness of the fit, the
smoothness factor \(k\) is restricted to a value of 3 (Wood, 2006). This
restriction was applied to smooth variables throughout this paper and we
have dropped this from the subsequent equations.

For the model in equation 2, we only used the data from Zhang et al.
(2017) to make sure that the additional watersheds added to the data set
did not influence the analysis. Given that in Zhang et al. (2017),
dryness (\(\frac{E0}{Pa}\)) is used to look at variations in the change
in flow, we also fitted the following model:

\[\tag{3}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) +  {forest~type} + \\  &{climate~type} + {assessment~type} + {hydrologic~regime} + \varepsilon
\end{aligned}\]

Subsequently, using the full data set, including the additional
watersheds and the additional variables the following two models were
fitted:

\[\tag{4}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(Pa) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
\end{aligned}\]

\[\tag{5}
\begin{aligned}
log10(\Delta Q) \sim ~&log10(\Delta forest) + s(\frac{E0}{Pa}) + s(Area) + s(Latitude) + s(Longitude) + \\
& s(years) + {forest~type} + {climate~type} + {assessment~type} + \\ & {hydrologic~regime} + \varepsilon
\end{aligned}\]

The results were analysed to identify:\\
1. the significance of the different variables\\
2. the direction of the categorical or shape of the smooth variables

@@

\leavevmode\hypertarget{ref-borg1988}{}%
Borg, H., Bell, R.W., Loh, I.C., 1988. Streamflow and stream salinity in
a small water supply catchment in southwest western australia after
reforestation. Journal of Hydrology 103, 323--333.
doi:\href{https://doi.org/https://doi.org/10.1016/0022-1694(88)90141-2}{https://doi.org/10.1016/0022-1694(88)90141-2}

@@

\leavevmode\hypertarget{ref-pena-arancibia2012}{}%
Peña-Arancibia, J.L., Dijk, A.I.J.M. van, Guerschman, J.P., Mulligan,
M., Bruijnzeel, L.A., McVicar, T.R., 2012. Detecting changes in
streamflow after partial woodland clearing in two large catchments in
the seasonal tropics. Journal of Hydrology 416-417, 60--71.
doi:\href{https://doi.org/https://doi.org/10.1016/j.jhydrol.2011.11.036}{https://doi.org/10.1016/j.jhydrol.2011.11.036}

\leavevmode\hypertarget{ref-rodriguez2010}{}%
Rodriguez, D.A., Tomasella, J., Linhares, C., 2010. Is the forest
conversion to pasture affecting the hydrological response of amazonian
catchments? Signals in the ji-paraná basin. Hydrological Processes 24,
1254--1269.
doi:\href{https://doi.org/https://doi.org/10.1002/hyp.7586}{https://doi.org/10.1002/hyp.7586}

\leavevmode\hypertarget{ref-ruprechtetal1991}{}%
Ruprecht, J.K., Schofield, N.J., Crombie, D.S., Vertessy, R.A.,
Stoneman, G.L., 1991. Early hydrological response to intense forest
thinning in southwestern australia. Journal of Hydrology 127, 261--277.
doi:\href{https://doi.org/https://doi.org/10.1016/0022-1694(91)90118-2}{https://doi.org/10.1016/0022-1694(91)90118-2}

\leavevmode\hypertarget{ref-thornton2007}{}%
Thornton, C.M., Cowie, B.A., Freebairn, D.M., Playford, C.L., 2007. The
brigalow catchment study: II*. Clearing brigalow (acacia harpophylla)
for cropping or pasture increases runoff. Australian Journal of Soil
Research 45, 496--511.
doi:\href{https://doi.org/doi:10.1071/SR07064}{doi:10.1071/SR07064}

@@

\leavevmode\hypertarget{ref-zhao2010}{}%
Zhao, F., Zhang, L., Xu, Z., Scott, D.F., 2010. Evaluation of methods
for estimating the effects of vegetation change and climate variability
on streamflow. Water Resources Research 46, W03505.
doi:\href{https://doi.org/10.1029/2009wr007702}{10.1029/2009wr007702}

\leavevmode\hypertarget{ref-zhou2010}{}%
Zhou, G., Wei, X., Luo, Y., Zhang, M., Li, Y., Qiao, Y., Liu, H., Wang,
C., 2010. Forest recovery and river discharge at the regional scale of
guangdong province, china. Water Resources Research 46.
doi:\href{https://doi.org/https://doi.org/10.1029/2009WR008829}{https://doi.org/10.1029/2009WR008829}
